---
title: "Lung function in children with asthma: a descriptive study"
subtitle: "Demonstration manuscript for the A Quarto Odyssey course"
author:
  - name: "[Put your name here]"
    email: "example@med.up.pt"
    affiliations:
      - "Faculty of Medicine, University of Porto"
date: today
date-format: "D MMMM YYYY"
abstract: |
  This document is the skeleton you'll complete in exercises 2 and 3 of
  the *A Quarto Odyssey* course. In exercise 2, you'll fill in the three R
  code blocks marked with `TODO`. In exercise 3, you'll write the
  Introduction and Methods sections, insert citations from your Zotero
  library, and export the document to .docx, PDF and HTML.
bibliography: references.bib
csl: vancouver.csl
format:
  html:
    toc: true
    code-fold: false
  docx: default
  pdf:
    documentclass: scrartcl
    geometry: margin=2.5cm
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup

# TODO: chunk 1 — load the tidyverse and gt, and read the file
# ../data/cohort-asma.csv into an object called `cohort`.
#
# Hint:
#   library(tidyverse)
#   library(gt)
#   cohort <- read_csv("../data/cohort-asma.csv")

```

# Introduction

> *In exercise 3, write two or three paragraphs here about paediatric asthma,
> with at least three citations from your Zotero library.*
>
> Citation syntax:
>
> - `@papi2018asthma` → "Papi et al. (2018)"
> - `[@papi2018asthma]` → "(Papi et al., 2018)" or "(1)" in Vancouver
> - `[@papi2018asthma; @quanjer2012multi]` → combines citations

# Methods

> *In exercise 3, write the sample description here. Include:*
>
> - *Sample size (n) and age range.*
> - *Distribution by sex.*
> - *Inclusion and exclusion criteria (made up — it's an exercise).*
> - *At least one reference to the statistical method used.*
>
> *Cross-reference the tables and figures with `@tbl-demographics` and `@fig-fev1-age`.*
>
> *Bonus (optional, see exercise 3): generate a STROBE flow diagram of the*
> *cohort with the `quarto-study-flow` extension and cross-reference it with `@fig-flow`.*

# Results

The sample is characterised in @tbl-demographics. The relationship between age and lung function, stratified by asthma severity, is illustrated in @fig-fev1-age.

```{r}
#| label: tbl-demographics
#| tbl-cap: "Demographic and functional characteristics of the sample, by sex"

# TODO: chunk 2 — produce a table with:
#   - n by sex
#   - mean (standard deviation) of age, by sex
#   - mean (standard deviation) of FEV1, by sex
#
# Hint (dplyr + gt):
#   cohort |>
#     group_by(sex) |>
#     summarise(
#       n = n(),
#       age_mean = mean(age),
#       age_sd = sd(age),
#       fev1_mean = mean(fev1_l),
#       fev1_sd = sd(fev1_l)
#     ) |>
#     gt() |>
#     fmt_number(columns = age_mean:fev1_sd, decimals = 2)

```

```{r}
#| label: fig-fev1-age
#| fig-cap: "FEV1 as a function of age, stratified by asthma severity"
#| fig-width: 7
#| fig-height: 5

# TODO: chunk 3 — ggplot2 plot with:
#   - x axis: age
#   - y axis: fev1_l
#   - colour: asthma_severity
#   - geom_point() + geom_smooth(method = "lm")
#
# Hint:
#   ggplot(cohort, aes(x = age, y = fev1_l, colour = asthma_severity)) +
#     geom_point(alpha = 0.7) +
#     geom_smooth(method = "lm", se = FALSE) +
#     labs(x = "Age (years)", y = "FEV1 (L)", colour = "Severity") +
#     theme_minimal()

```

# Discussion

> *You don't need to write the discussion. The goal of the exercise is the
> Quarto workflow, not the clinical analysis.*

# References

::: {#refs}
:::
