---
title: "Lung function in children with asthma: a descriptive study"
subtitle: "Demonstration manuscript for the A Quarto Odyssey course"
author:
  - name: "[Put your name here]"
    email: "example@med.up.pt"
    affiliations:
      - "Faculty of Medicine, University of Porto"
date: today
date-format: "D MMMM YYYY"
abstract: |
  This is a ready-made manuscript: the R code already works and the text is
  already written. Your job is the *Quarto layer*, not the analysis. In
  exercise 2 you configure the code chunks (hide the code, add labels and
  captions, size the figure) and add cross-references. In exercise 3 you
  format the text, insert citations from your Zotero library, and export the
  document to .docx, PDF and HTML.
bibliography: references.bib
csl: vancouver.csl
format:
  html:
    toc: true
    code-fold: false
  docx: default
  pdf:
    documentclass: scrartcl
    geometry: margin=2.5cm
# NOTE (exercise 2, step 2): there is no `execute:` block yet. Render the
# document once and you'll see the R code and its messages in the output.
# You'll add the block here to hide them.
---

```{r}
# This setup chunk loads the packages and reads the data. It works as-is —
# don't change the code. In exercise 2 you'll configure how it appears.
library(tidyverse)
library(gt)

cohort <- read_csv("../data/cohort-asma.csv")
```

# Introduction

<!-- TODO (exercise 3): this paragraph is written, but flat. Format it and add
     the citations at the [CITE] marks using your Zotero cite keys, e.g.
     [@papi2018asthma]. -->

Asthma is one of the most common chronic diseases of childhood [CITE]. Spirometry — in particular the forced expiratory volume in one second (FEV1) — is central to assessing and monitoring disease severity in children [CITE]. Reference values for paediatric spirometry follow widely adopted multi-ethnic equations [CITE].

# Methods

<!-- TODO (exercise 3): format this section (a short bulleted list of the sample
     characteristics reads well here) and add the cross-references and a citation
     for the statistical approach. -->

We analysed a synthetic cohort of children with asthma. For each participant we recorded age, sex, height, weight and spirometry (FEV1, FVC). The sample is summarised with descriptive statistics — means and standard deviations by sex [CITE]. Sample characteristics are reported in TABLE-REF and the relationship between age and lung function in FIGURE-REF.

# Results

The demographic and functional characteristics of the sample are shown in the table below. The relationship between age and FEV1, stratified by asthma severity, is shown in the figure below.

<!-- TODO (exercise 2): once you give the two chunks below a label, replace
     "the table below" / "the figure below" above with cross-references,
     e.g. @tbl-demographics and @fig-fev1-age. -->

```{r}
# TODO (exercise 2): configure this chunk — give it a label (tbl-demographics)
# and a caption (tbl-cap). The code is complete; don't change it.
cohort |>
  group_by(sex) |>
  summarise(
    n = n(),
    age_mean = mean(age),
    age_sd = sd(age),
    fev1_mean = mean(fev1_l),
    fev1_sd = sd(fev1_l)
  ) |>
  gt() |>
  fmt_number(columns = age_mean:fev1_sd, decimals = 2) |>
  cols_label(
    sex = "Sex",
    n = "N",
    age_mean = "Age (mean)",
    age_sd = "Age (SD)",
    fev1_mean = "FEV1 (mean)",
    fev1_sd = "FEV1 (SD)"
  )
```

```{r}
# TODO (exercise 2): configure this chunk — give it a label (fig-fev1-age),
# a caption (fig-cap), and set fig-width / fig-height. The code is complete.
ggplot(cohort, aes(x = age, y = fev1_l, colour = asthma_severity)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Age (years)", y = "FEV1 (L)", colour = "Severity") +
  theme_minimal()
```

# Discussion

<!-- The discussion is intentionally left short — the goal of the exercise is the
     Quarto workflow, not the clinical interpretation. -->

These descriptive results illustrate the expected increase of FEV1 with age. A full analysis is beyond the scope of this demonstration.

# References

::: {#refs}
:::
